---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# matchatr

<!-- badges: start -->
[![R-CMD-check](https://github.com/etverse/matchatr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/etverse/matchatr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**matchatr** provides causal inference for **(matched) case-control, nested
case-control (NCC), and case-cohort** study designs. It pairs design-faithful
classical estimators with marginal causal effects, and integrates with the
[etverse](https://github.com/etverse) ecosystem — delegating estimation to
[`causatr`](https://github.com/etverse/causatr) (g-computation / IPW / AIPW with
sandwich and bootstrap variance) and [`survatr`](https://github.com/etverse/survatr)
(causal survival on person-period data).

> **Status: classical engines complete; first marginal causal contrast landed.**
> The design taxonomy, the two-step `matcha()` / `contrast()` API, and the
> `(design, estimator)` dispatch (`PHASE_1`) are in place, and the classical
> engines run end to end: the **unmatched** case-control logistic and
> Mantel-Haenszel ORs (`PHASE_2`), the **matched** case-control
> conditional-logistic and McNemar ORs with stratum-specific effect modification
> (`PHASE_3`), the **polytomous** subtype ORs with `test_homogeneity()`
> (`PHASE_4`), the **nested case-control** risk-set hazard ratio (`PHASE_5`), the
> **case-cohort** Prentice / Self-Prentice / Borgan hazard ratios and absolute
> risk (`PHASE_6`), and the **Samuelsen IPW** weighted Cox / AFT / additive-hazards
> analyses of NCC data with absolute and excess risk (`PHASE_7`). The **causal
> layer** is now opening: `estimator = "ccw_gformula"` / `"ccw_ipw"` / `"ccw_aipw"`
> report a **marginal** risk difference, risk ratio, or marginal odds ratio from a
> case-control sample with a known prevalence q₀, via Rose & van der Laan
> case-control weighting delegated to `causatr` g-computation / IPW / (doubly-robust)
> AIPW (`PHASE_9` Chunks 1–2). See the
> [articles](https://etverse.github.io/matchatr/) for worked examples. The
> remaining case-control-weighting estimators (IPW / AIPW / TMLE), the
> design-weighted causal survival layer, and the efficiency phases (`PHASE_9`
> Chunks 2–4, `PHASE_10`–`PHASE_20`) remain at the design stage.

## What it does

Two orthogonal axes: a **`design`** object encodes the sampling structure (strata,
matching ratio, time scale, prevalence, inclusion weights); an **`estimator`** chooses
the analysis.

| Design | Classical estimand | Causal (marginal) estimand |
|---|---|---|
| Unmatched case-control | conditional OR, Mantel-Haenszel | RD / RR / marginal OR (case-control weighting) |
| Matched case-control | conditional OR (conditional logistic) | RD / RR via standardization |
| Nested case-control | risk-set HR; Samuelsen IPW Cox | marginal survival contrasts (design-weighted) |
| Case-cohort | Prentice / Self-Prentice / Borgan HR | absolute risk, RD(t), RMST |

Marginal causal effects use **case-control weighting** (the Rose & van der Laan
g-formula / IPW / AIPW / TMLE family) and **design-based inclusion weighting**
(Samuelsen, Borgan): the weights are passed as observation weights into the etverse
engines, so they compose directly with existing estimators.

## Installation

You can install the development version of matchatr from
[GitHub](https://github.com/etverse/matchatr) with:

``` r
# install.packages("pak")
pak::pak("etverse/matchatr")
```

## Example

```{r example}
library(matchatr)

# Matched case-control -> conditional odds ratio (infert: a matched study of
# spontaneous/induced abortion and infertility, matched on age and parity).
fit <- matcha(
  infert,
  outcome = "case", exposure = "spontaneous",
  design = matched_cc(strata = "stratum"),
  confounders = ~ induced, estimator = "clogit"
)

contrast(fit, type = "or")
```

The marginal causal contrasts (case-control weighting) reuse the same two-step
API once a source-population prevalence q0 is supplied — they are part of the
roadmap below:

```{r example-causal, eval = FALSE}
fit <- matcha(
  data,
  outcome = "case", exposure = "x",
  design = unmatched_cc(prevalence = 0.02),   # source-population prevalence q0
  confounders = ~ age + smoke, estimator = "ccw_gformula"
)
contrast(fit, type = "difference", ci_method = "sandwich")
```

## Roadmap

The design is documented in `PHASE_1`–`PHASE_20` at the repository root, mapping the
*Handbook of Statistical Methods for Case-Control Studies* (Borgan et al., 2018) to an
implementation plan. See `CLAUDE.md` for the phase index and `FEATURE_COVERAGE_MATRIX.md`
for what is implemented and tested.

## Part of the etverse

matchatr is one package in the [etverse](https://github.com/etverse) family for causal
inference and methodological triangulation, alongside `causatr` (causal effect
estimation) and `survatr` (causal survival analysis).
