---
title: "Case-cohort hazard ratios and absolute risk"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Case-cohort hazard ratios and absolute risk}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
nocite: |
  @self1988asymptotic
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A *case-cohort* study draws a random **subcohort** from the full cohort at
baseline and augments it with every case (regardless of subcohort membership).
The key advantage over nested case-control sampling is that the subcohort serves
as a control group simultaneously for all endpoints of interest — even endpoints
ascertained years apart — because selection happens once, at the start. Controls
are reused across failure times, which is what makes the pseudo-likelihood
analysis necessary.

matchatr wraps `survival::cch()`, which implements the Prentice, Self-Prentice,
Lin-Ying, and Borgan I/II pseudo-likelihoods with their correct asymptotic
variances. The additional absolute-risk layer (IPW Breslow cumulative baseline
hazard) is matchatr's own.

```{r}
#| message: false
library(matchatr)
library(survival)
```

## The Wilms tumour data

We use `survival::nwtco`, the standard case-cohort example from the
*Handbook of Statistical Methods for Case-Control Studies* [@borgan2018handbook].
A random subcohort
of 668 subjects was drawn from 4 028 children enrolled in the National Wilms
Tumour Study. The endpoint is relapse (`rel`); the exposure of interest is
histological type (`histol`: 1 = unfavourable). The subcohort indicator is
`in.subcohort`.

```{r}
nwtco2 <- nwtco
nwtco2$subcohort <- as.logical(nwtco2$in.subcohort)
nwtco2$stage     <- factor(nwtco2$stage)

cat("Cohort size: ", nrow(nwtco2), "\n")
cat("Subcohort:   ", sum(nwtco2$subcohort), " (~",
    round(100 * mean(nwtco2$subcohort)), "%)\n", sep = "")
cat("Relapses:    ", sum(nwtco2$rel), "\n")
```

`matcha()` takes the **full cohort** — it subsets internally to
cases + subcohort members before calling `cch()`, so the denominators for the
per-stratum cohort sizes are computed correctly.

## Hazard ratio: Prentice pseudo-likelihood

The @prentice1986casecohort pseudo-likelihood augments each risk set with the case if
the case falls outside the subcohort. `contrast()` reports the hazard ratio with
the Self-Prentice corrected asymptotic variance (the naive information-matrix SE
is not valid here because controls are reused across failure times).

```{r}
fit <- matcha(
  nwtco2,
  outcome     = "rel",
  exposure    = "histol",
  design      = case_cohort(subcohort = "subcohort", time = "edrel",
                            method = "Prentice"),
  confounders = ~ stage + age,
  estimator   = "cch"
)
fit
contrast(fit)       # type = "hr" is the default for a case-cohort design
```

Unfavourable histology roughly doubles the relapse hazard conditional on stage
and age at diagnosis.

## Alternative pseudo-likelihoods

Three further methods are supported. `"SelfPrentice"` uses the subcohort alone
at each failure time (slightly more efficient); `"LinYing"` uses the full
subcohort plus all failures [@lin1993cox].

```{r}
fit_sp <- matcha(nwtco2, "rel", "histol",
                 design = case_cohort("subcohort", "edrel",
                                      method = "SelfPrentice"),
                 confounders = ~ stage + age, estimator = "cch")
fit_ly <- matcha(nwtco2, "rel", "histol",
                 design = case_cohort("subcohort", "edrel",
                                      method = "LinYing"),
                 confounders = ~ stage + age, estimator = "cch")

rbind(
  Prentice    = unlist(contrast(fit)$contrasts[1, c("estimate", "ci_lower", "ci_upper")]),
  SelfPrentice = unlist(contrast(fit_sp)$contrasts[1, c("estimate", "ci_lower", "ci_upper")]),
  LinYing     = unlist(contrast(fit_ly)$contrasts[1, c("estimate", "ci_lower", "ci_upper")])
)
```

All three methods agree closely; differences reflect the distinct variance
estimators rather than the point estimates.

## Stratified subcohorts: Borgan I/II

When the subcohort is drawn by stratified sampling (e.g. oversampling a rare
exposure group), the @borgan2000exposure IPW estimators weight each risk-set
contribution by the inverse of its stratum-specific subcohort sampling fraction
`N_s / n_sub_s`. Provide the stratification column via `stratum =`.

```{r}
fit_b2 <- matcha(
  nwtco2,
  outcome     = "rel",
  exposure    = "histol",
  design      = case_cohort(subcohort = "subcohort", time = "edrel",
                            method = "II.Borgan", stratum = "stage"),
  confounders = ~ age,
  estimator   = "cch"
)
contrast(fit_b2)
```

Requesting a Borgan method without supplying `stratum` is declined:

```{r}
#| error: true
matcha(nwtco2, "rel", "histol",
       design = case_cohort("subcohort", "edrel", method = "I.Borgan"),
       confounders = ~ stage + age, estimator = "cch")
```

## Absolute risk: IPW Breslow cumulative baseline hazard

`absolute_risk()` evaluates the cumulative incidence
F_x(t) = 1 − exp(−exp(β̂ᵀ x) Λ̂₀(t)) at specified times, where Λ̂₀(t) is the
IPW Breslow estimator. The subcohort-only risk-set denominator at each event
time is scaled by the inverse sampling fraction `N / n_sub` (or per-stratum
`N_s / n_sub_s` for Borgan methods) so the estimate targets the full-cohort
baseline hazard. Pointwise confidence intervals use the delta method on the
complementary log-log scale [@borgan1990note].

```{r}
# Two covariate profiles: favourable vs unfavourable histology, median age
nd <- data.frame(
  histol = c(0L, 1L),
  stage  = factor(c(2, 2), levels = 1:4),
  age    = c(3, 3)
)

ar <- absolute_risk(fit, newdata = nd, times = c(500, 1000, 2000))
tidy(ar)
```

We can compare the case-cohort absolute risk against the full-cohort Kaplan-Meier
/ Cox curves. Because the subcohort is only ~17% of the cohort, individual-run
sampling variability is visible, but the estimates track the full-cohort curve
closely.

```{r}
cox_full <- coxph(Surv(edrel, rel) ~ histol + stage + age, data = nwtco2)
sf_full  <- survfit(cox_full, newdata = nd)
f_full   <- summary(sf_full, times = c(500, 1000, 2000))$surv
f_full   <- 1 - f_full

cat("Full-cohort F_x(t) [histol=0, histol=1]:\n")
print(round(f_full, 3))

cat("\nIPW Breslow F_x(t) [histol=0, histol=1]:\n")
est_wide <- matrix(ar$estimates$estimate, nrow = 2, byrow = FALSE)
print(round(est_wide, 3))
```

## One scale per design

A case-cohort fit identifies the **hazard ratio**. Requesting an odds ratio
(appropriate for matched case-control data) is declined:

```{r}
#| error: true
contrast(fit, type = "or")
```

A marginal risk difference / risk ratio from the *classical* case-cohort analysis
would require a source-population prevalence `q0` that `estimator = "cch"` does not
supply — but the design-weighted g-computation below gets there a different way.

## Marginal causal survival

The hazard ratio above is a *conditional* effect on the hazard scale. For a
**marginal** causal contrast — the absolute risk, risk difference, risk ratio, or
restricted mean survival time (RMST) the whole cohort would show under one exposure
versus the other — use `estimator = "surv_gcomp"`. It g-computes on the same
design-weighted Cox: it standardizes the model-based absolute risk
$F(t \mid x, W) = 1 - \exp(-\hat\Lambda_0(t)\,e^{\hat\beta^\top x})$ over the
subcohort (a random sample of the cohort, weighted by the inverse subcohort sampling
fraction) to the treat-all and treat-none marginal risks, then contrasts them. No
`q0` is needed: a case-cohort *samples the cohort directly*, so the subcohort already
represents the source-population covariate distribution. Variance is a
design-preserving bootstrap.

The exposure must be binary, so we code unfavourable histology as a 0/1 indicator:

```{r}
nwtco2$histol_uh <- as.integer(nwtco2$histol == 2L)

fit_g <- matcha(
  nwtco2,
  outcome     = "rel",
  exposure    = "histol_uh",
  design      = case_cohort(subcohort = "subcohort", time = "edrel"),
  confounders = ~ stage + age,
  estimator   = "surv_gcomp"
)

# Marginal risk difference at two relapse-free follow-up times (days).
contrast(fit_g, type = "difference", times = c(1500, 3000), n_boot = 200)
```

Unfavourable histology raises the marginal relapse risk by roughly 28 percentage
points — a population-level statement the hazard ratio cannot make directly. The
restricted mean survival time difference up to a horizon summarises the gap as
expected relapse-free days lost:

```{r}
contrast(fit_g, type = "rmst", times = 3000, n_boot = 200)
```

The risk ratio (`type = "ratio"`) is available on the same fit.

## Covered combinations

**Legend.** ✅ truth-pinned in tests · ⛔ rejected with an informative error.

| Method | Subcohort | Estimand | Variance | Status |
|---|---|---|---|---|
| Prentice | simple | HR | Self-Prentice asymptotic | ✅ |
| SelfPrentice / LinYing | simple | HR | asymptotic | ✅ |
| I.Borgan | stratified | HR | Borgan asymptotic | ✅ |
| II.Borgan | stratified | HR | plug-in asymptotic | ✅ |
| any method | — | absolute risk F_x(t) | IPW Breslow + delta log-log CI | ✅ |
| surv_gcomp | simple / stratified | marginal RD(t) / RR(t) / RMST | design-preserving bootstrap | ✅ |
| Borgan I/II, no `stratum` | — | — | ⛔ `matchatr_bad_design` |
| cch | — | odds ratio | ⛔ use `type = "hr"` |
| cch | — | RD / RR | ⛔ need `q0` (use `surv_gcomp`) |
| cch | — | sandwich / bootstrap CI | ⛔ `matchatr_unsupported_variance` |
| surv_gcomp | — | conditional OR / HR | ⛔ `matchatr_unidentified_estimand` |
| surv_gcomp | — | non-binary exposure | ⛔ `matchatr_bad_input` |

See `FEATURE_COVERAGE_MATRIX.md` for the full combination table.

## References

::: {#refs}
:::
