---
title: "Marginal causal effects from a case-control sample"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Marginal causal effects from a case-control sample}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every estimator in the previous articles reports a **conditional odds ratio** (or
a conditional hazard ratio). That is a perfectly good *within-stratum* effect, but
it is not what a population decision usually needs:

- it is **non-collapsible** — it changes when you add or drop a covariate, even
  with no confounding, so the "adjusted OR" is not a marginal effect;
- it is on the **odds-ratio scale** only — there is no risk difference or risk
  ratio, and no absolute risk, because a case-control sample fixes the outcome
  margin by design and so carries no information about baseline risk.

A **marginal** causal effect — the average treatment effect as a risk difference
(RD), risk ratio (RR), or marginal odds ratio (mOR) — needs one extra ingredient:
the source-population outcome prevalence **q₀ = P(Y = 1)**. With q₀ we can undo the
case-oversampling and recover the marginal effect by **case-control weighting**
(Rose & van der Laan, 2008).

```{r}
#| message: false
library(matchatr)
```

## The idea: reweight the sample back to the population

A case-control study oversamples cases. If we weight each case by
q₀ / (sample case fraction) and each control by (1 − q₀) / (sample control
fraction), the *weighted* sample has outcome margin q₀ again — its weighted
empirical distribution mimics the source cohort. A cohort estimator applied to the
weighted sample (here g-computation: fit a weighted outcome model and standardize
over the covariates) then targets the **marginal** estimand. matchatr builds these
weights internally; you only supply q₀ on the design via
`unmatched_cc(prevalence = q0)`.

The point estimate and its variance are delegated to the etverse causal engine
[`causatr`](https://github.com/etverse/causatr); matchatr owns the weighting and
sampling-design layer.

## A cohort with a known marginal effect

We simulate a cohort with a binary exposure `x` confounded by a continuous `w`, so
the conditional and marginal effects genuinely differ. Because we know the true
outcome model, we can compute the marginal truth analytically by the g-formula:
average the counterfactual risks under "everyone exposed" and "everyone
unexposed" over the cohort.

```{r}
set.seed(13)
n      <- 1e5
alpha  <- -3
beta_x <- log(2.5)        # the CONDITIONAL log odds ratio for x
gamma  <- 1.0
w  <- rnorm(n)
x  <- rbinom(n, 1, plogis(0.2 * w))                 # x confounded by w
y  <- rbinom(n, 1, plogis(alpha + beta_x * x + gamma * w))

# g-formula truth on the full cohort (true model): mean counterfactual risks.
m1 <- mean(plogis(alpha + beta_x + gamma * w))      # treat-all
m0 <- mean(plogis(alpha + gamma * w))               # treat-none
truth <- c(
  RD  = m1 - m0,
  RR  = m1 / m0,
  mOR = (m1 / (1 - m1)) / (m0 / (1 - m0)),
  conditional_OR = exp(beta_x)
)
round(truth, 4)
```

Note the marginal OR (`mOR`) is **not** the conditional OR `exp(beta_x) = 2.5`:
non-collapsibility pulls the marginal OR toward the null.

Now draw an unmatched case-control sample — every case plus a 5:1 random sample of
controls — and record the source prevalence q₀ = P(Y = 1).

```{r}
q0 <- mean(y)
cohort   <- data.frame(case = y, x = x, w = w)
cases    <- cohort[cohort$case == 1, ]
controls <- cohort[cohort$case == 0, ]
controls <- controls[sample.int(nrow(controls), 5 * nrow(cases)), ]
cc <- rbind(cases, controls)
rownames(cc) <- NULL
c(q0 = round(q0, 4), n_cases = nrow(cases), n_controls = nrow(controls))
```

## The marginal effect via case-control weighting

`matcha()` with `estimator = "ccw_gformula"` and a design carrying q₀ fits the
case-control-weighted g-formula. `contrast()` then reports the marginal effect on
the requested scale — the **risk difference is the default**.

```{r}
fit <- matcha(
  cc,
  outcome = "case", exposure = "x",
  design = unmatched_cc(prevalence = q0),
  confounders = ~ w,
  estimator = "ccw_gformula"
)

contrast(fit, type = "difference")   # marginal risk difference (default)
contrast(fit, type = "ratio")        # marginal risk ratio
contrast(fit, type = "or")           # marginal odds ratio
```

Each estimate recovers its analytical marginal truth:

```{r}
est <- c(
  RD  = contrast(fit, type = "difference")$contrasts$estimate,
  RR  = contrast(fit, type = "ratio")$contrasts$estimate,
  mOR = contrast(fit, type = "or")$contrasts$estimate
)
round(rbind(estimate = est, truth = truth[c("RD", "RR", "mOR")]), 4)
```

## Marginal is not conditional

The same sample analysed with the plain conditional logistic estimator returns the
**conditional** odds ratio `exp(beta_x) = 2.5` — a different number from the
marginal OR above. This is the non-collapsibility distinction that motivates
case-control weighting: the two estimators answer different questions.

```{r}
cond <- matcha(
  cc,
  outcome = "case", exposure = "x",
  design = unmatched_cc(),     # no q0: only the conditional OR is identified
  confounders = ~ w,
  estimator = "logistic"
)
c(
  conditional_OR = contrast(cond, type = "or")$contrasts$estimate,
  marginal_OR    = contrast(fit,  type = "or")$contrasts$estimate
)
```

## What is declined

Case-control weighting needs q₀ and targets a marginal effect, so a few requests
are declined with informative, classed errors:

```{r}
#| error: true
# No q0 on the design: the sample cannot be reweighted to the population.
matcha(
  cc, outcome = "case", exposure = "x",
  design = unmatched_cc(), estimator = "ccw_gformula"
)
```

```{r}
#| error: true
# The CCW g-formula reports a marginal effect, not a hazard ratio.
contrast(fit, type = "hr")
```

```{r}
#| error: true
# A bootstrap interval must resample within case / control strata and recompute
# the q0 weights each replicate (a later chunk); only the sandwich is wired here.
contrast(fit, type = "difference", ci_method = "bootstrap")
```

A non-binary exposure is also declined: this estimator reports the binary average
treatment effect (treat-all versus treat-none), so a continuous or multi-level
exposure is routed to a conditional estimator instead.

## Three estimators, one weighting concept

The same q₀ weights drive a whole family of cohort estimators on the weighted
sample. `estimator = "ccw_ipw"` weights by an inverse-propensity model instead of
standardizing an outcome model, and `estimator = "ccw_aipw"` combines both into the
**augmented** (doubly-robust) estimator. All three report the same marginal effect
when their models are correct:

```{r}
ests <- c("ccw_gformula", "ccw_ipw", "ccw_aipw")
rd <- sapply(ests, function(e) {
  f <- matcha(cc, outcome = "case", exposure = "x",
              design = unmatched_cc(prevalence = q0),
              confounders = ~ w, estimator = e)
  contrast(f, type = "difference")$contrasts$estimate
})
round(rbind(marginal_RD = rd, truth = truth["RD"]), 4)
```

## Double robustness of CCW-AIPW

CCW-AIPW's appeal is that it is consistent if **either** the outcome model **or**
the propensity model is correct — you get two chances to be right. We can see this
directly. Build a cohort where the outcome truly depends on `w` non-linearly but
the propensity is linear in `w`; fitting both working models as `~ w` (linear)
then misspecifies the **outcome** model while the **propensity** model is correct:

```{r}
set.seed(20)
n <- 1e5; w <- rnorm(n)
a <- rbinom(n, 1, plogis(2 * w))                          # propensity linear in w: CORRECT
lp <- function(av) -2.3 + 0.7 * av + 2 * w - 2.5 * w^2    # outcome non-linear: a ~w model is WRONG
y <- rbinom(n, 1, plogis(lp(a)))
truth_rd <- mean(plogis(lp(1)) - plogis(lp(0)))           # g-formula truth
q0b <- mean(y)
co <- data.frame(case = y, x = a, w = w)
ca <- co[co$case == 1, ]; ct <- co[co$case == 0, ]
ct <- ct[sample.int(nrow(ct), 5 * nrow(ca)), ]
ccb <- rbind(ca, ct); rownames(ccb) <- NULL

rd2 <- sapply(ests, function(e) {
  f <- matcha(ccb, outcome = "case", exposure = "x",
              design = unmatched_cc(prevalence = q0b),
              confounders = ~ w, estimator = e)
  contrast(f, type = "difference")$contrasts$estimate
})
round(rbind(marginal_RD = rd2, truth = truth_rd), 4)
```

With the outcome model misspecified, `ccw_gformula` is biased, but `ccw_ipw` (correct
propensity) and `ccw_aipw` recover the truth. Swap the misspecification to the
propensity model and it is `ccw_ipw` that fails while `ccw_gformula` and `ccw_aipw`
stay on target — AIPW is robust to either.

## Where this sits

The case-control-weighting family now offers g-formula, IPW, and the doubly-robust
AIPW marginal contrasts for an unmatched case-control study with a known prevalence.
Still to come (forthcoming chunks): CCW-TMLE (a targeting step), the estimated-q₀
variance correction, a design-respecting bootstrap, and matched / nested
case-control support — all built on the same one weighting concept.
