---
title: "Introduction to matchatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Introduction to matchatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

matchatr is the etverse package for causal inference from **case-control-type
study designs**: (matched) case-control, nested case-control, and case-cohort
samples. It provides design-faithful *classical* estimators (conditional
logistic regression, Mantel-Haenszel, McNemar, weighted Cox, case-cohort
pseudo-likelihood) and *marginal causal* effects via case-control weighting,
delegating the heavy lifting to its siblings
[`causatr`](https://github.com/etverse/causatr) (g-computation / IPW / AIPW) and
[`survatr`](https://github.com/etverse/survatr) (causal survival).

This vignette introduces the design taxonomy and the two-step API. The
estimator-specific articles cover the worked examples:

- [Unmatched case-control](unmatched-cc.qmd) — logistic and Mantel-Haenszel
  odds ratios.
- [Matched case-control](matched-cc.qmd) — conditional logistic, McNemar, and
  stratum-specific effect modification.
- [Multiple groups](multiple-groups.qmd) — polytomous (multinomial) subtype
  odds ratios.

## Two orthogonal axes: design and estimator

Every matchatr analysis is specified along two independent axes:

- A **`design`** object encodes the *sampling structure* — how the sample was
  drawn from the source population. It carries the strata / matched-set ids, the
  time scale (for risk-set designs), the source-population prevalence q0, and the
  intended weighting scheme.
- An **`estimator`** chooses the *analysis* — conditional vs marginal, odds
  ratio vs hazard ratio vs risk difference.

The two are deliberately separate: the same matched case-control sample can be
analysed by conditional logistic regression (a conditional odds ratio) or, with
a prevalence q0, by a case-control-weighted g-formula (a marginal risk
difference). You change the estimand without re-describing the design.

```{r}
#| message: false
library(matchatr)

# Six design constructors, one per sampling structure:
unmatched_cc()                              # independent case / control sampling
matched_cc(strata = "set", ratio = 2)       # individually / frequency matched
nested_cc(strata = "set", time = "t")       # risk-set sampling from a cohort
```

Each returns a `matchatr_design` object that prints its structure:

```{r}
matched_cc(strata = c("age_grp", "sex"), ratio = 2)
```

The full set of constructors is `unmatched_cc()`, `matched_cc()`,
`nested_cc()`, `case_cohort()`, `two_phase()`, and `counter_matched()`.

## The two-step API

matchatr mirrors the etverse verb convention (`causatr::causat()`,
`survatr::surv_fit()`):

```r
# Step 1 — fit: resolve the (design, estimator) pair and run the engine.
fit <- matcha(data, outcome = "case", exposure = "x",
              design = unmatched_cc(), confounders = ~ age + smoke,
              estimator = "logistic")

# Step 2 — contrast: report the effect on the requested scale.
contrast(fit, type = "or")
```

`matcha()` returns a `matchatr_fit`; `contrast()` returns a `matchatr_result`.
When you omit `estimator`, the design's canonical default is used; when you omit
`type`, `contrast()` reports the estimand the design identifies.

## A worked example

We use R's built-in `infert` data — a matched case-control study of prior
spontaneous / induced abortion and secondary infertility, matched on age and
parity (the `stratum` column identifies the matched sets).

```{r}
fit <- matcha(
  infert,
  outcome = "case", exposure = "spontaneous",
  design = matched_cc(strata = "stratum"),
  confounders = ~ induced, estimator = "clogit"
)
fit
```

The fit echoes the resolved engine and the case / control counts. The second
step reports the exposure's conditional odds ratio:

```{r}
contrast(fit, type = "or")
```

Each prior spontaneous abortion multiplies the conditional odds of infertility
by about 7, adjusting for induced abortions and the matched design.

## Inspecting a fit

A `matchatr_fit` works with the broom-style generics. `tidy()` returns the
coefficient table (log-odds scale, or the odds-ratio scale with
`exponentiate = TRUE`); `summary()` prints the model summary.

```{r}
tidy(fit, exponentiate = TRUE)
```

`contrast()` results also tidy to a one-row-per-contrast table:

```{r}
tidy(contrast(fit, type = "or"))
```

## What is identified depends on the design

From a case-control sample the marginal outcome frequency is fixed by the
sampling, so only the **odds ratio** is identified without extra information.
Asking for a risk difference or risk ratio aborts with an informative,
classed error pointing to the prevalence q0 you would need:

```{r}
#| error: true
contrast(fit, type = "difference")
```

Supplying the source-population prevalence on the design
(`unmatched_cc(prevalence = 0.02)`) unlocks the case-control-weighted marginal
contrasts — that layer is on the roadmap below.

## What works today

matchatr is built phase by phase against the *Handbook of Statistical Methods
for Case-Control Studies* (Borgan et al., 2018). The classical odds-ratio
engines are implemented:

| Design | Estimator | Estimand | Article |
|---|---|---|---|
| Unmatched case-control | `"logistic"` | conditional OR (any exposure type) | [Unmatched CC](unmatched-cc.qmd) |
| Unmatched case-control | `"mh"` | Mantel-Haenszel stratified OR | [Unmatched CC](unmatched-cc.qmd) |
| Matched case-control | `"clogit"` | conditional OR (+ effect modification) | [Matched CC](matched-cc.qmd) |
| Matched case-control | `"mcnemar"` | 1:1 matched-pair OR | [Matched CC](matched-cc.qmd) |
| Multi-group outcome | `"polytomous"` | per-subtype OR vs reference | [Multiple groups](multiple-groups.qmd) |

The time-to-event sampling designs (nested case-control, case-cohort, IPW-NCC)
and the marginal causal layer (case-control-weighting g-formula / IPW / AIPW /
TMLE, design-weighted causal survival) are designed but not yet implemented; see
`FEATURE_COVERAGE_MATRIX.md` for the authoritative status of every cell.

## References

Borgan Ø, Breslow N, Chatterjee N, Gail MH, Scott A, Wild CJ (2018). *Handbook
of Statistical Methods for Case-Control Studies*. Chapman & Hall/CRC.

Rose S, van der Laan MJ (2009). Why match? Investigating matched case-control
study designs with causal effect estimation. *The International Journal of
Biostatistics* 5(1).
