---
title: "IPW analysis of nested case-control data"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{IPW analysis of nested case-control data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The **classical** nested case-control (NCC) analysis conditions on the sampled
risk sets: each case and its matched controls form a stratum, and
`estimator = "clogit"` uses the conditional partial likelihood (see
`vignette("nested-cc")`). Controls appear only at their matched case's failure
time — a design feature that removes residual confounding by the matching
variables but also discards information each control carries across the full
follow-up.

The **inverse probability weighting (IPW)** approach *breaks the matching*. Each
sampled control receives a weight equal to the inverse of its inclusion
probability under the NCC sampling scheme, then a standard weighted Cox partial
likelihood is fit over the unique subjects (cases + unique controls). Controls
now contribute at every event time where they are still at risk — their exposure
information is re-used across all failure times rather than only at their matched
event. This gains efficiency, especially when many controls were matched to early
cases and their follow-up extends across later events.

```{r}
#| message: false
library(matchatr)
```

## Generating a cohort and an NCC sample

```{r}
set.seed(77)
n      <- 3000
beta_x <- log(2)                       # true Cox log hazard ratio
exposure   <- rbinom(n, 1, 0.4)
confounder <- rnorm(n)
rate       <- 0.08 * exp(beta_x * exposure + 0.4 * confounder)
tt         <- rexp(n, rate)            # single draw for consistent t and d
cohort <- data.frame(
  id         = seq_len(n),
  t          = pmin(tt, 5),
  d          = as.integer(tt <= 5),
  exposure   = exposure,
  confounder = confounder
)
```

`sample_ncc()` with `incl_prob = TRUE` draws the NCC sample and appends two
columns needed for the IPW analysis:

-   **`ipw_weight`** — the Samuelsen (1997) inverse inclusion probability
    1/π_j for each sampled control; cases get weight 1. The inclusion
    probability is the Kaplan-Meier-type formula
    π_j = 1 − prod of (1 − m_i / n_elig_i) over all event times where j
    was eligible, where m_i is the controls sampled and n_elig_i the eligible
    pool size.
-   **`.cohort_row`** — original cohort row index, used to deduplicate controls
    that were sampled into multiple risk sets before fitting the Cox model.

```{r}
set.seed(1)
ncc <- sample_ncc(cohort, time = "t", event = "d", m = 3, incl_prob = TRUE)
ncc[1:6, c("id", "t", "d", "exposure", "case", "set", "risk_time",
           "ipw_weight", ".cohort_row")]
```

## Fitting the IPW weighted Cox model

With `estimator = "ipw_cox"`, matchatr:

1.  Deduplicates the NCC data by `.cohort_row` (each cohort subject appears
    once in the analysis).
2.  Fits `coxph(Surv(t, d) ~ exposure + confounder, weights = ipw_weight,
    robust = TRUE)`.
3.  Returns the Lin-Wei robust sandwich variance via `coxph(robust = TRUE)`.

```{r}
fit_ipw <- matcha(
  ncc,
  outcome     = "d",          # cohort event indicator, not the per-set "case"
  exposure    = "exposure",
  design      = nested_cc(strata = "set", time = "t"),
  confounders = ~confounder,
  estimator   = "ipw_cox"
)
fit_ipw
```

```{r}
contrast(fit_ipw)             # reports type = "hr" by default
```

## Comparison: classical NCC vs IPW NCC vs full cohort

The conditional partial likelihood (`clogit`) and the IPW weighted Cox target
the same hazard ratio; the IPW estimator is at least as efficient because it
reuses controls across failure times.

```{r}
# Classical NCC (clogit, matched analysis) — same seed as the IPW sample above
set.seed(1)
ncc_classical <- sample_ncc(cohort, time = "t", event = "d", m = 3)
fit_clogit <- matcha(ncc_classical, "case", "exposure",
                     nested_cc(strata = "set", time = "t"),
                     confounders = ~confounder, estimator = "clogit")

# Full-cohort Cox (the gold standard)
fit_full <- survival::coxph(
  survival::Surv(t, d) ~ exposure + confounder, data = cohort
)

cat("Full-cohort Cox HR: ", round(exp(coef(fit_full)["exposure"]), 3), "\n")
cat("Classical NCC HR:   ", round(contrast(fit_clogit)$contrasts$estimate, 3), "\n")
cat("IPW NCC HR:         ", round(contrast(fit_ipw)$contrasts$estimate, 3), "\n")
```

All three estimates target the same hazard ratio; the classical and IPW NCC
estimates will differ by a small simulation noise term.

## tidy() output

```{r}
tidy(contrast(fit_ipw))
```

## Rejection paths

```{r}
#| error: true
# Requesting an odds ratio from an IPW NCC fit is rejected (wrong scale).
contrast(fit_ipw, type = "or")
```

```{r}
#| error: true
# Bootstrap is not available for the IPW Cox estimator.
contrast(fit_ipw, type = "hr", ci_method = "bootstrap")
```

```{r}
#| error: true
# The ipw_weight column is required; sample_ncc() without incl_prob = TRUE
# produces data that lacks it.
ncc_no_wts <- sample_ncc(cohort, "t", "d", m = 3)
matcha(ncc_no_wts, "case", "exposure",
       nested_cc(strata = "set", time = "t"),
       estimator = "ipw_cox")
```

## References

Samuelsen SO (1997). A pseudolikelihood approach to analysis of nested
case-control studies. *Biometrika*, 84(2), 379–394.

Stoer NC & Samuelsen SO (2013). Inverse probability weighting in nested
case-control studies with additional matching — a simulation study.
*Statistics in Medicine*, 32(30), 5328–5339.
