---
title: "IPW analysis of nested case-control data"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{IPW analysis of nested case-control data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The **classical** nested case-control (NCC) analysis conditions on the sampled
risk sets: each case and its matched controls form a stratum, and
`estimator = "clogit"` uses the conditional partial likelihood (see
`vignette("nested-cc")`). Controls appear only at their matched case's failure
time — a design feature that removes residual confounding by the matching
variables but also discards information each control carries across the full
follow-up.

The **inverse probability weighting (IPW)** approach *breaks the matching*. Each
sampled control receives a weight equal to the inverse of its inclusion
probability under the NCC sampling scheme, then a standard weighted Cox partial
likelihood is fit over the unique subjects (cases + unique controls). Controls
now contribute at every event time where they are still at risk — their exposure
information is re-used across all failure times rather than only at their matched
event. This gains efficiency, especially when many controls were matched to early
cases and their follow-up extends across later events.

```{r}
#| message: false
library(matchatr)
```

## Generating a cohort and an NCC sample

```{r}
set.seed(77)
n      <- 3000
beta_x <- log(2)                       # true Cox log hazard ratio
exposure   <- rbinom(n, 1, 0.4)
confounder <- rnorm(n)
rate       <- 0.08 * exp(beta_x * exposure + 0.4 * confounder)
tt         <- rexp(n, rate)            # single draw for consistent t and d
cohort <- data.frame(
  id         = seq_len(n),
  t          = pmin(tt, 5),
  d          = as.integer(tt <= 5),
  exposure   = exposure,
  confounder = confounder
)
```

`sample_ncc()` with `incl_prob = TRUE` draws the NCC sample and appends two
columns needed for the IPW analysis:

-   **`ipw_weight`** — the Samuelsen (1997) inverse inclusion probability
    1/π_j for each sampled control; cases get weight 1. The inclusion
    probability is the Kaplan-Meier-type formula
    π_j = 1 − prod of (1 − m_i / n_elig_i) over all event times where j
    was eligible, where m_i is the controls sampled and n_elig_i the eligible
    pool size.
-   **`.cohort_row`** — original cohort row index, used to deduplicate controls
    that were sampled into multiple risk sets before fitting the Cox model.

```{r}
set.seed(1)
ncc <- sample_ncc(cohort, time = "t", event = "d", m = 3, incl_prob = TRUE)
ncc[1:6, c("id", "t", "d", "exposure", "case", "set", "risk_time",
           "ipw_weight", ".cohort_row")]
```

## Fitting the IPW weighted Cox model

With `estimator = "ipw_cox"`, matchatr:

1.  Deduplicates the NCC data by `.cohort_row` (each cohort subject appears
    once in the analysis).
2.  Fits `coxph(Surv(t, d) ~ exposure + confounder, weights = ipw_weight,
    robust = TRUE)`.
3.  Returns the Lin-Wei robust sandwich variance via `coxph(robust = TRUE)`.

```{r}
fit_ipw <- matcha(
  ncc,
  outcome     = "d",          # cohort event indicator, not the per-set "case"
  exposure    = "exposure",
  design      = nested_cc(strata = "set", time = "t"),
  confounders = ~confounder,
  estimator   = "ipw_cox"
)
fit_ipw
```

```{r}
contrast(fit_ipw)             # reports type = "hr" by default
```

## Comparison: classical NCC vs IPW NCC vs full cohort

The conditional partial likelihood (`clogit`) and the IPW weighted Cox target
the same hazard ratio; the IPW estimator is at least as efficient because it
reuses controls across failure times.

```{r}
# Classical NCC (clogit, matched analysis) — same seed as the IPW sample above
set.seed(1)
ncc_classical <- sample_ncc(cohort, time = "t", event = "d", m = 3)
fit_clogit <- matcha(ncc_classical, "case", "exposure",
                     nested_cc(strata = "set", time = "t"),
                     confounders = ~confounder, estimator = "clogit")

# Full-cohort Cox (the gold standard)
fit_full <- survival::coxph(
  survival::Surv(t, d) ~ exposure + confounder, data = cohort
)

cat("Full-cohort Cox HR: ", round(exp(coef(fit_full)["exposure"]), 3), "\n")
cat("Classical NCC HR:   ", round(contrast(fit_clogit)$contrasts$estimate, 3), "\n")
cat("IPW NCC HR:         ", round(contrast(fit_ipw)$contrasts$estimate, 3), "\n")
```

All three estimates target the same hazard ratio; the classical and IPW NCC
estimates will differ by a small simulation noise term.

## tidy() output

```{r}
tidy(contrast(fit_ipw))
```

## Rejection paths

```{r}
#| error: true
# Requesting an odds ratio from an IPW NCC fit is rejected (wrong scale).
contrast(fit_ipw, type = "or")
```

```{r}
#| error: true
# Bootstrap is not available for the IPW Cox estimator.
contrast(fit_ipw, type = "hr", ci_method = "bootstrap")
```

```{r}
#| error: true
# The ipw_weight column is required; sample_ncc() without incl_prob = TRUE
# produces data that lacks it.
ncc_no_wts <- sample_ncc(cohort, "t", "d", m = 3)
matcha(ncc_no_wts, "case", "exposure",
       nested_cc(strata = "set", time = "t"),
       estimator = "ipw_cox")
```

## Working-model weights

Samuelsen KM weights are **design-based**: they depend only on the risk-set
structure and are computed during `sample_ncc()`. An alternative is to fit a
**working model** for the selection probability — a logistic regression (or GAM)
of whether a subject was selected as a control, given its covariates and the
event time. Working-model weights can be more efficient when the logistic model
is close to the truth.

`compute_ncc_weights()` replaces the `ipw_weight` column with working-model
estimates. It requires the **full Phase-1 cohort** to reconstruct who was
eligible (but not selected) at each event time.

```{r}
# Replace KM weights with GLM working-model weights
ncc_glm <- compute_ncc_weights(
  ncc,
  cohort           = cohort,
  method           = "glm",
  selection_formula = ~ risk_time,   # default: time-only logistic model
  time             = "t"
)
ncc_glm[1:3, c("id", "case", "set", "ipw_weight")]
```

Fit and compare:

```{r}
fit_glm <- matcha(
  ncc_glm,
  outcome     = "d",
  exposure    = "exposure",
  design      = nested_cc(strata = "set", time = "t"),
  confounders = ~confounder,
  estimator   = "ipw_cox"
)
cat("KM weights HR: ", round(contrast(fit_ipw)$contrasts$estimate, 3), "\n")
cat("GLM weights HR:", round(contrast(fit_glm)$contrasts$estimate, 3), "\n")
```

Both KM and GLM weights target the same Cox hazard ratio; differences reflect
the working-model vs design-based weight formulas.

A GAM can be used when the selection probability is non-linearly related to time:

```{r}
#| warning: false
ncc_gam <- compute_ncc_weights(
  ncc,
  cohort           = cohort,
  method           = "gam",
  selection_formula = ~ s(risk_time),
  time             = "t"
)
fit_gam <- matcha(
  ncc_gam,
  outcome     = "d",
  exposure    = "exposure",
  design      = nested_cc(strata = "set", time = "t"),
  confounders = ~confounder,
  estimator   = "ipw_cox"
)
cat("GAM weights HR:", round(contrast(fit_gam)$contrasts$estimate, 3), "\n")
```

The Phase-1 cohort is required: omitting it aborts with `matchatr_missing_phase1`.

```{r}
#| error: true
compute_ncc_weights(ncc, cohort = NULL, method = "glm", time = "t")
```

## Absolute risk

Because the IPW reformulation builds a *weighted pseudo-cohort*, it recovers the
cumulative baseline hazard — which the classical conditional partial likelihood
conditions away. `absolute_risk()` turns the fitted weighted Cox into a
cumulative incidence

$$F_x(t) = 1 - \exp\!\big(-\exp(\hat\beta^{\top} x)\,\hat\Lambda_0(t)\big),$$

where $\hat\Lambda_0(t)$ is an inverse-probability-weighted Breslow cumulative
baseline hazard over the deduplicated analysis sample: at each event time the
at-risk denominator $\sum_j w_j \exp(\hat\beta^{\top} x_j)$ upweights each control
by $1/\pi_j$ so it stands in for the unsampled cohort. Confidence intervals use
the delta method on the complementary log-log scale.

```{r}
newdata <- data.frame(exposure = c(1, 0), confounder = c(0, 0))
ar <- absolute_risk(fit_ipw, newdata = newdata, times = c(1, 2, 3, 4))
ar
```

The hand-rolled weighted Breslow agrees with `survival::survfit` on the same
weighted Cox model to machine precision, so the point estimate is exact; the
interval is conservative (the within-sample Nelson-Aalen baseline variance).

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-alt: >
#|   Cumulative incidence over follow-up for an exposed versus an unexposed
#|   subject estimated from the IPW nested case-control weighted Cox model. The
#|   exposed curve rises faster, reflecting the hazard ratio above one.
est <- ar$estimates
exposed   <- est[est$row == 1, ]
unexposed <- est[est$row == 2, ]
plot(exposed$time, exposed$estimate, type = "b", pch = 19, col = "#c0392b",
     ylim = c(0, max(est$ci_upper)), xlab = "time", ylab = expression(hat(F)[x](t)),
     main = "IPW NCC cumulative incidence")
lines(unexposed$time, unexposed$estimate, type = "b", pch = 19, col = "#2980b9")
legend("topleft", c("exposed", "unexposed"), col = c("#c0392b", "#2980b9"),
       pch = 19, lty = 1, bty = "n")
```

The same verb serves the case-cohort `cch` engine (see `vignette("case-cohort")`);
only the baseline-hazard weighting differs.

## Multiple endpoints from one control set

The real payoff of breaking the matching is **control reuse across endpoints**.
The classical conditional analysis ties each control to its case's failure time,
so a control set sampled for one endpoint cannot serve another. Under IPW the
sample is a weighted pseudo-cohort, so the *same* controls can fit a weighted Cox
for any endpoint recorded in the cohort. matchatr supports two modes.

We build a cohort with two competing endpoints — a harmful exposure for endpoint
1, a protective one for endpoint 2:

```{r}
set.seed(20)
n <- 4000
x <- rbinom(n, 1, 0.4)
z <- rnorm(n)
t1 <- rexp(n, 0.05 * exp(log(2.0) * x + 0.3 * z))   # endpoint 1: HR 2.0
t2 <- rexp(n, 0.05 * exp(log(0.5) * x + 0.2 * z))   # endpoint 2: HR 0.5
tau <- 6
tt    <- pmin(t1, t2, tau)
cause <- ifelse(tt >= tau, 0L, ifelse(t1 < t2, 1L, 2L))
co <- data.frame(
  id = seq_len(n), t = tt,
  d_any = as.integer(cause != 0L),   # any failure
  d1    = as.integer(cause == 1L),   # endpoint 1
  d2    = as.integer(cause == 2L),   # endpoint 2
  x = x, z = z
)
```

### Mode A — sample on the union event, analyse each endpoint

Sampling on the "any-failure" event ascertains *every* endpoint's cases at once,
so each cause-specific endpoint is analysed directly by passing it as `outcome`.
The single sampled control set, with its shared inclusion weights, backs both
fits:

```{r}
set.seed(1)
ncc_multi <- sample_ncc(co, time = "t", event = "d_any", m = 3, incl_prob = TRUE)

fit_e1 <- matcha(ncc_multi, outcome = "d1", exposure = "x",
                 design = nested_cc(strata = "set", time = "t"),
                 confounders = ~ z, estimator = "ipw_cox")
fit_e2 <- matcha(ncc_multi, outcome = "d2", exposure = "x",
                 design = nested_cc(strata = "set", time = "t"),
                 confounders = ~ z, estimator = "ipw_cox")

rbind(cbind(endpoint = "d1", tidy(contrast(fit_e1))),
      cbind(endpoint = "d2", tidy(contrast(fit_e2))))
```

Both hazard ratios sit near their cohort truths (2.0 and 0.5). A competing-endpoint
case is *ascertained* by the sampling, so it keeps weight 1 (not the control
weight 1/π_j) when a different endpoint is analysed — matchatr handles this when
it builds the deduplicated analysis sample.

### Mode B — reuse a primary-endpoint control set for a secondary endpoint

If the NCC was drawn for a single *primary* endpoint, its controls can still be
reused for a *secondary* one. `reuse_ncc_endpoint()` keeps the controls' primary
inclusion weights 1/π_j (the inclusion probability is a property of the sampling,
not the endpoint) and augments the secondary endpoint's cases that were not
sampled — pulled from the Phase-1 cohort with weight 1, since every case of an
analysed endpoint is ascertained with probability 1:

```{r}
set.seed(1)
ncc_primary <- sample_ncc(co, time = "t", event = "d1", m = 3, incl_prob = TRUE)

ncc_reuse <- reuse_ncc_endpoint(ncc_primary, cohort = co, time = "t", event = "d2")
fit_reuse <- matcha(ncc_reuse, outcome = "d2", exposure = "x",
                    design = nested_cc(strata = "set", time = "t"),
                    confounders = ~ z, estimator = "ipw_cox")
tidy(contrast(fit_reuse))
```

Reusing the secondary endpoint needs the cohort to recover its unsampled cases;
omitting it is rejected:

```{r}
#| error: true
reuse_ncc_endpoint(ncc_primary, cohort = NULL, time = "t", event = "d2")
```

## Alternative models: time ratios and excess hazards

Breaking the matching also frees the analysis from the proportional-hazards
scale. The same Samuelsen-weighted sample supports two non-Cox models, each
reporting a different effect measure.

The **accelerated failure time** model (`estimator = "ipw_aft"`) acts on the
survival-*time* scale: `contrast(type = "af")` reports the **time ratio**
exp(β) — the factor by which a unit of exposure multiplies the survival time
(above one prolongs, below one shortens). It is a weighted Weibull `survreg`
with a robust sandwich.

```{r}
fit_aft <- matcha(ncc, outcome = "d", exposure = "exposure",
                  design = nested_cc(strata = "set", time = "t"),
                  confounders = ~ confounder, estimator = "ipw_aft")
contrast(fit_aft)
```

The fitted Weibull is a full *survival curve*, so `absolute_risk()` also serves
the `ipw_aft` engine — reading the cumulative incidence straight off the
parameters, with no Breslow step:

$$F_x(t) = 1 - \exp\!\big(-\exp((\log t - \hat\eta)/\hat\sigma)\big),$$

where $\hat\eta = \hat\beta^{\top} x$ is the AFT linear predictor and
$\hat\sigma$ the scale. The interval again uses the delta method on the
complementary log-log scale, here over $(\beta, \log\sigma)$ with the robust
`survreg` sandwich. Because the curve is parametric, $\hat F_x(t)$ is smooth in
$t$ (the Cox engine's $\hat F_x(t)$ is a step function on the event times).

```{r}
ar_aft <- absolute_risk(fit_aft, newdata = newdata, times = c(1, 2, 3, 4))
ar_aft
```

The baseline defaults to Weibull, but `dist` selects any of four AFT
distributions — `"weibull"`, `"exponential"`, `"lognormal"`, `"loglogistic"`.
All report the same time ratio `exp(β)`; they differ in the survival-curve shape
(`"lognormal"` / `"loglogistic"` allow a non-monotone hazard). `absolute_risk()`
follows the chosen baseline automatically:

```{r}
fit_ln <- matcha(ncc, outcome = "d", exposure = "exposure",
                 design = nested_cc(strata = "set", time = "t"),
                 confounders = ~ confounder, estimator = "ipw_aft",
                 dist = "lognormal")
contrast(fit_ln)
```

The **additive-hazards** model (`estimator = "ipw_aalen"`) acts on the *rate*
scale: it writes the hazard as a sum, λ(t) = λ₀(t) + γ·exposure, so
`contrast(type = "excess")` reports the **excess hazard** γ — an additive rate
difference. Unlike a ratio it can be negative (a protective exposure lowers the
rate), and its confidence interval is symmetric (it is not exponentiated).

```{r}
fit_add <- matcha(ncc, outcome = "d", exposure = "exposure",
                  design = nested_cc(strata = "set", time = "t"),
                  confounders = ~ confounder, estimator = "ipw_aalen")
contrast(fit_add)
```

Each estimator identifies exactly one scale, so asking for another is rejected:

```{r}
#| error: true
contrast(fit_aft, type = "hr")
```

## References

Lin DY & Ying Z (1994). Semiparametric analysis of the additive risk model.
*Biometrika*, 81(1), 61–71.

Borgan Ø & Langholz B (1997). Estimation of excess risk from case-control data
using Aalen's linear regression model. *Biometrics*, 53(2), 690–697.

Samuelsen SO (1997). A pseudolikelihood approach to analysis of nested
case-control studies. *Biometrika*, 84(2), 379–394.

Saarela O, Kulathinal S, Arjas E & Läärä E (2008). Nested case–control data
utilized for multiple outcomes: a likelihood approach and alternatives.
*Statistics in Medicine*, 27(28), 5991–6008.

Støer NC & Samuelsen SO (2012). Comparison of estimators in nested case–control
studies with multiple outcomes. *Lifetime Data Analysis*, 18(3), 261–283.

Støer NC & Samuelsen SO (2013). Inverse probability weighting in nested
case-control studies with additional matching — a simulation study.
*Statistics in Medicine*, 32(30), 5328–5339.

Kang S, Lu W & Liu M (2017). Efficient estimation for accelerated failure time
model under case-cohort and nested case-control sampling. *Biometrics*, 73(1),
114–123.
