---
title: "Matched case-control odds ratios"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Matched case-control odds ratios}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In a *matched* case-control study each case is matched to one or more controls
sharing the values of the matching variable(s). The design-faithful analysis is
**conditional logistic regression** (conditional maximum likelihood): it
conditions on the matched-set totals, which removes the matching-variable
nuisance parameters. Fitting matched-set indicators by ordinary logistic
regression instead biases the odds ratio — for 1:1 matching its estimate
converges to the *squared* OR (Breslow & Day, 1980) — so matchatr never does
that.

matchatr provides three tools on `matched_cc()`:

- `estimator = "clogit"` — the conditional logistic OR for any matched-set
  composition (1:1, m:1, variable-ratio).
- `estimator = "mcnemar"` — the closed-form 1:1 matched-pair OR.
- `effect_modifier = ` — stratum-specific conditional ORs from an
  `exposure × modifier` interaction.

```{r}
#| message: false
library(matchatr)
```

## Conditional logistic regression

We use the built-in `infert` data: a matched case-control study of secondary
infertility and prior abortion, with cases matched to controls on age and parity
(`stratum` identifies the matched sets).

```{r}
fit <- matcha(
  infert,
  outcome = "case", exposure = "spontaneous",
  design = matched_cc(strata = "stratum"),
  confounders = ~ induced, estimator = "clogit"
)
fit
```

Conditioning on the matched-set totals controls the matching variables (age,
parity) implicitly, so only the exposure and adjustment ORs are reported:

```{r}
contrast(fit, type = "or")
```

The interval is the partial-likelihood information-matrix Wald interval.
`"clogit"` is the design's default estimator, so `estimator = "clogit"` may be
omitted. m:1 and variable-ratio matching need no special handling — the
conditional likelihood treats any matched-set composition uniformly.

## McNemar's 1:1 matched-pair OR

For genuine 1:1 pairs there is a closed form: with `n10` pairs where only the
case is exposed and `n01` where only the control is exposed, the matched-pair
odds ratio is `OR = n10 / n01` with `Var(log OR) = 1/n10 + 1/n01` (McNemar,
1947). Pairs concordant on exposure carry no information and cancel.

We simulate matched pairs from the conditional likelihood with a known
conditional OR of 3:

```{r}
set.seed(4)
n_pairs <- 1200
OR_true <- 3
# Each pair is exposure-discordant with probability 0.4; among discordant pairs
# the case (rather than the control) is the exposed member with probability
# OR/(1 + OR), so n10/n01 estimates the conditional OR.
discordant <- rbinom(n_pairs, 1, 0.4)
p_case_exposed <- OR_true / (1 + OR_true)
case_exposed <- ifelse(discordant == 1, rbinom(n_pairs, 1, p_case_exposed),
                       rbinom(n_pairs, 1, 0.5))
ctrl_exposed <- ifelse(discordant == 1, 1L - case_exposed, case_exposed)

pairs <- data.frame(
  set    = rep(seq_len(n_pairs), each = 2),
  case   = rep(c(1L, 0L), times = n_pairs),
  smoker = as.vector(rbind(case_exposed, ctrl_exposed))
)

fit_mn <- matcha(
  pairs,
  outcome = "case", exposure = "smoker",
  design = matched_cc(strata = "set"), estimator = "mcnemar"
)
contrast(fit_mn, type = "or")
```

This is exact, not a model fit. And because the conditional likelihood *reduces*
to McNemar's for 1:1 binary data, `clogit` on the same pairs gives an identical
point estimate and interval:

```{r}
fit_cl <- matcha(
  pairs,
  outcome = "case", exposure = "smoker",
  design = matched_cc(strata = "set"), estimator = "clogit"
)
contrast(fit_cl, type = "or")
```

`"mcnemar"` applies only to genuine 1:1 pairs with a binary exposure: a richer
matched set (m:1) is rejected with a pointer to `"clogit"`, as is a non-binary
exposure.

## Stratum-specific effect modification

To ask whether the exposure effect *differs* across the levels of a categorical
modifier, pass `effect_modifier =`. The conditional logistic engine then fits
`outcome ~ exposure * modifier + confounders + strata(set)`, and
`contrast(type = "or")` reports the exposure's conditional OR *within each
modifier level* — the main coefficient at the reference level and the linear
combination β_x + β_{x:level} elsewhere — with Wald intervals from the joint
partial-likelihood variance (so each per-level SE accounts for the
main–interaction covariance).

Here we split `infert` by maternal age and ask whether the spontaneous-abortion
OR differs between younger and older women:

```{r}
infert2 <- infert
infert2$age_grp <- factor(ifelse(infert2$age < 31, "younger", "older"))

fit_em <- matcha(
  infert2,
  outcome = "case", exposure = "spontaneous",
  design = matched_cc(strata = "stratum"), estimator = "clogit",
  effect_modifier = "age_grp"
)
contrast(fit_em, type = "or")
```

The modifier may coincide with a matching variable — that is the canonical use
(does the exposure effect vary across the matching factor?). Effect modification
is supported for a single-coefficient exposure (binary, continuous, or two-level
factor) crossed with a categorical modifier; a 3+-level factor exposure or a
numeric modifier is rejected with a classed error.

## What the design does and does not identify

As with the unmatched design, only the odds ratio is identified; risk
differences and risk ratios need a source-population prevalence q0 and a
case-control-weighted estimator. The conditional fit reports the
information-matrix interval only, so `ci_method = "sandwich"` / `"bootstrap"`
are declined for these engines.

## Covered combinations

**Legend.** ✅ truth-pinned in tests · ⛔ rejected with an informative error.

| Matching | Estimator | Estimand | Status |
|---|---|---|---|
| 1:1 / m:1 / variable-ratio | `clogit` | conditional OR | ✅ |
| 1:1 (binary exposure) | `mcnemar` | matched-pair OR | ✅ |
| any | `clogit` + `effect_modifier` | stratum-specific OR | ✅ |
| m:1 (or richer) | `mcnemar` | — | ⛔ use `clogit` |
| non-binary exposure | `mcnemar` | — | ⛔ use `clogit` |
| any | `clogit` / `mcnemar` | RD / RR | ⛔ need q0 |

See `FEATURE_COVERAGE_MATRIX.md` for the authoritative status of every
combination.

## References

Breslow NE, Day NE (1980). *Statistical Methods in Cancer Research, Volume 1:
The Analysis of Case-Control Studies*. IARC.

McNemar Q (1947). Note on the sampling error of the difference between
correlated proportions or percentages. *Psychometrika* 12(2):153-157.

Pike MC, Hill AP, Smith PG (1980). Bias and efficiency in logistic analyses of
stratified case-control studies. *International Journal of Epidemiology*
9(1):89-95.
