---
title: "Multiple case / control groups"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Multiple case / control groups}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Some case-control studies have more than two outcome groups: several *disease
subtypes* (does the exposure act differently on each?), or several *control
groups*. The design-faithful analysis is **baseline-category multinomial
logistic regression**, fit by `nnet::multinom`. Each non-reference equation's
exposure coefficient is that subtype's log odds ratio versus the reference, so
`contrast(type = "or")` reports one OR per (subtype, exposure coefficient).

```{r}
#| message: false
library(matchatr)
```

## A three-group outcome

We simulate an outcome with a control group and two disease subtypes, where the
binary exposure raises the odds of subtype A by a factor of 2 and subtype B by a
factor of 4 (versus control).

```{r}
set.seed(5)
n <- 6000
x <- rbinom(n, 1, 0.4)
# Baseline-category logits relative to "control": OR_A = 2, OR_B = 4.
eta <- cbind(
  control = 0,
  A       = -1.0 + log(2) * x,
  B       = -1.5 + log(4) * x
)
prob <- exp(eta) / rowSums(exp(eta))
subtype <- apply(prob, 1, function(p) sample(c("control", "A", "B"), 1, prob = p))
d_poly <- data.frame(subtype = subtype, x = x)
table(d_poly$subtype)
```

Select `estimator = "polytomous"` and name the baseline group with `reference`.
The outcome must be a factor or character column with three or more groups (a
two-group, numeric, or logical outcome is routed back to the binary estimators).

```{r}
fit <- matcha(
  d_poly,
  outcome = "subtype", exposure = "x",
  design = unmatched_cc(),
  estimator = "polytomous", reference = "control"
)
fit
```

The fit prints the per-group counts and flags the reference group. The contrast
reports each subtype's exposure OR versus the reference, recovering the true 2
and 4:

```{r}
contrast(fit, type = "or")
```

## The full per-equation table

`tidy()` renders the whole multinomial model, one equation per non-reference
group, with a leading `y.level` column naming the group each row contrasts
against the reference:

```{r}
tidy(fit, exponentiate = TRUE)
```

The interval is the multinomial information matrix; the Huber-White sandwich is
not defined for this fit, so `robust = TRUE` / `ci_method = "sandwich"` are
declined.

## Choosing the reference and adjusting for confounders

The `reference` group is releveled to the front before fitting, so it can be any
observed group. Confounders enter the same way as in the binary engines, holding
the exposure ORs adjusted:

```{r}
set.seed(6)
d_poly$age <- rnorm(n, 50, 10)

fit_adj <- matcha(
  d_poly,
  outcome = "subtype", exposure = "x",
  design = unmatched_cc(),
  confounders = ~ age,
  estimator = "polytomous", reference = "control"
)
contrast(fit_adj, type = "or")
```

## A note on collinearity

Unlike `stats::glm`, `nnet::multinom` does not alias a rank-deficient exposure
column to `NA` — it splits the coefficient across the outcome equations, which
would silently attenuate the OR. matchatr guards against this by checking the
design-matrix rank up front: an exposure that is constant, or collinear with a
confounder, is rejected with a classed `matchatr_unestimable_exposure` error
rather than returning a misleading number. Collinearity confined to the
confounders leaves the exposure estimable and is not rejected.

## Covered combinations

**Legend.** ✅ truth-pinned in tests · ⛔ rejected with an informative error.

| Outcome | Estimator | Estimand | Status |
|---|---|---|---|
| ≥3-group factor / character | `polytomous` | per-subtype OR vs reference | ✅ |
| ≥3-group (confounder-adjusted) | `polytomous` | adjusted per-subtype OR | ✅ |
| 2-group / numeric / logical | `polytomous` | — | ⛔ use `logistic` / `clogit` |
| matched design | `polytomous` | — | ⛔ unsupported |
| any | `polytomous` | RD / RR | ⛔ need q0 |
| any | `polytomous` | sandwich / bootstrap variance | ⛔ information matrix only |

See `FEATURE_COVERAGE_MATRIX.md` for the authoritative status of every
combination.

## References

Borgan Ø, Breslow N, Chatterjee N, Gail MH, Scott A, Wild CJ (2018). *Handbook
of Statistical Methods for Case-Control Studies*, Chapter 5. Chapman & Hall/CRC.

Venables WN, Ripley BD (2002). *Modern Applied Statistics with S*, 4th ed.
Springer. (The `nnet::multinom` fitter.)
