---
title: "Nested case-control hazard ratios"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Nested case-control hazard ratios}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A *nested* case-control (NCC) study samples controls from inside a cohort by
**risk-set (incidence-density) sampling**: at each case's failure time, a few
controls are drawn at random from the subjects still at risk. The design-faithful
analysis is the **conditional partial likelihood** with each sampled risk set as
a stratum — exactly the conditional logistic regression matchatr already uses for
matched case-control data, because a matched set and a sampled risk set are the
same stratum construction.

The estimand, however, is different. Under proper risk-set sampling the
conditional estimate is the **hazard ratio**, exactly — there is no rare-disease
approximation involved (Prentice & Breslow, 1978). So matchatr reports it on the
hazard-ratio scale (`type = "hr"`), which is the default for `nested_cc()`.

```{r}
#| message: false
library(matchatr)
```

## A cohort and a nested sample

We simulate a cohort with a constant baseline hazard and a known Cox log hazard
ratio for a binary exposure `x`.

```{r}
set.seed(51)
n <- 3000
x <- rbinom(n, 1, 0.4)
beta <- log(2.2)                       # the true log hazard ratio for x
rate <- 0.08 * exp(beta * x)           # PH hazard: baseline times exp(lp)
time <- rexp(n, rate)
tau  <- 4                              # administrative censoring
cohort <- data.frame(
  id = seq_len(n), t = pmin(time, tau), d = as.integer(time <= tau), x = x
)
```

`sample_ncc()` draws the risk-set (incidence-density) sample: for each case it
takes `m` controls at random from the subjects still at risk at that failure
time. The result is analysis-ready — it carries a per-set `case` indicator
(distinct from the cohort-wide `event`, since a sampled control may itself fail
later), a matched-set id `set`, and the set's `risk_time`.

```{r}
set.seed(1)
ncc <- sample_ncc(cohort, time = "t", event = "d", m = 3)
head(ncc)
```

Two further sampling options match real designs: `match = ~ sex + birth_cohort`
confines each case's controls to its own population stratum, and `entry =`
supplies a delayed-entry (left-truncation) column so a subject counts as at risk
only after it enters follow-up. A case left with no eligible control aborts with
an informative `matchatr_empty_risk_set` error rather than a degenerate set.

## The risk-set hazard ratio

`matcha()` with `design = nested_cc()` fits the conditional partial likelihood;
`contrast()` reports the hazard ratio with the partial-likelihood
information-matrix Wald interval.

```{r}
fit <- matcha(
  ncc,
  outcome = "case", exposure = "x",
  design = nested_cc(strata = "set", time = "risk_time"),
  estimator = "clogit"
)
fit

contrast(fit)        # type = "hr" is the default for a nested design
```

The estimate recovers the true log hazard ratio `log(2.2) ≈ 0.79`. The design's
`time` column records how the controls were sampled; the risk-set membership is
read from `strata`, so the conditional likelihood does not enter `time` here.

## OR = HR exactly

The whole point of risk-set sampling is that the sampled odds ratio *is* the
hazard ratio, with no rare-disease assumption. We can see this directly: the
NCC estimate matches the Cox hazard ratio fit on the **full cohort** (within
sampling error, since the NCC sample is a subset).

```{r}
full_cohort <- survival::coxph(survival::Surv(t, d) ~ x, data = cohort)
c(ncc_HR    = exp(coef(fit$model)[["x"]]),
  cohort_HR = exp(coef(full_cohort)[["x"]]))
```

## One scale per design

Each conditional design identifies exactly one scale: the matched design reports
the odds ratio, the nested design the hazard ratio. Asking for the off-design
scale is declined — the value would be the same number, but it is not the
estimand the design targets:

```{r}
#| error: true
contrast(fit, type = "or")
```

As with the other case-control designs, a marginal risk difference / risk ratio
needs a source-population prevalence `q0` and a case-control-weighted estimator,
and the conditional fit reports the information-matrix interval only, so
`ci_method = "sandwich"` / `"bootstrap"` are declined.

## Covered combinations

**Legend.** ✅ truth-pinned in tests · ⛔ rejected with an informative error.

| Sampling | Estimator | Estimand | Status |
|---|---|---|---|
| `sample_ncc()` risk-set draw (+ `match` / `entry`) | — | NCC sample from a cohort | ✅ |
| m:1 risk-set | `clogit` | conditional hazard ratio | ✅ |
| m:1 risk-set + confounder | `clogit` | conditional hazard ratio | ✅ |
| risk-set + `effect_modifier` | `clogit` | stratum-specific hazard ratio | ✅ |
| nested | `clogit` | odds ratio | ⛔ use `type = "hr"` |
| nested | `clogit` | RD / RR | ⛔ need q0 |
| nested, non-binary outcome | — | — | ⛔ |

See `FEATURE_COVERAGE_MATRIX.md` for the authoritative status of every
combination.

## References

Prentice RL, Breslow NE (1978). Retrospective studies and failure time models.
*Biometrika* 65(1):153-158.

Goldstein L, Langholz B (1992). Asymptotic theory for nested case-control
sampling in the Cox regression model. *Annals of Statistics* 20(4):1903-1928.

Thomas DC (1977). Addendum to "Methods of cohort analysis: appraisal by
application to asbestos mining" by Liddell FDK, McDonald JC, Thomas DC.
*Journal of the Royal Statistical Society A* 140(4):469-491.
