---
title: "Unmatched case-control odds ratios"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Unmatched case-control odds ratios}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In an *unmatched* case-control study, cases and controls are sampled
independently from the source population — no individual or frequency matching.
The classical analysis is the **conditional odds ratio**, identified from the
case-control sample even though the marginal disease frequency is fixed by the
sampling design. matchatr provides two engines on `unmatched_cc()`:

- `estimator = "logistic"` — logistic regression
  (`outcome ~ exposure + confounders`), the conditional OR for any exposure
  type. Because case-control sampling shifts only the intercept (Prentice &
  Pyke, 1979), the slope ORs equal the source-population ORs.
- `estimator = "mh"` — the closed-form Mantel-Haenszel stratified OR with a
  Robins-Breslow-Greenland confidence interval.

```{r}
#| message: false
library(matchatr)
```

## Logistic conditional OR

### A binary exposure

We simulate a cohort with a known log odds ratio, then draw an unmatched
case-control sample (all cases plus an equal number of controls). The conditional
OR should recover the cohort log-OR.

```{r}
set.seed(1)
n <- 8000
x <- rbinom(n, 1, 0.4)
age <- rnorm(n, 50, 10)
# True conditional log-OR for x is log(2.5).
lp <- -1 + log(2.5) * x + 0.03 * (age - 50)
case <- rbinom(n, 1, plogis(lp))
cohort <- data.frame(case = case, x = x, age = age)

# Unmatched case-control sample: every case + an equal random sample of controls.
cases <- cohort[cohort$case == 1, ]
controls <- cohort[cohort$case == 0, ]
cc <- rbind(cases, controls[sample(nrow(controls), nrow(cases)), ])

fit <- matcha(
  cc,
  outcome = "case", exposure = "x",
  design = unmatched_cc(),
  confounders = ~ age, estimator = "logistic"
)
contrast(fit, type = "or")
```

The recovered OR is close to the true 2.5 even though the sample is roughly 50%
cases — the odds ratio is invariant to the case / control sampling fractions.

The full coefficient table is available through `tidy()` (the intercept is
*not* an interpretable baseline risk under case-control sampling, so treat it as
a nuisance):

```{r}
tidy(fit, exponentiate = TRUE)
```

### A continuous exposure

A continuous exposure yields a single per-unit OR:

```{r}
set.seed(2)
n <- 6000
dose <- rnorm(n)
case <- rbinom(n, 1, plogis(-1 + 0.8 * dose))
d_cont <- data.frame(case = case, dose = dose)

fit_cont <- matcha(
  d_cont,
  outcome = "case", exposure = "dose",
  design = unmatched_cc(), estimator = "logistic"
)
contrast(fit_cont, type = "or")
```

### A categorical exposure: the `esoph` dose-response

The built-in `esoph` data is the Ille-et-Vilaine case-control study of
oesophageal cancer (Breslow & Day, 1980), stored as per-cell case / control
counts. We expand it to one row per subject so it can enter the logistic engine.

```{r}
esoph_long <- do.call(rbind, lapply(seq_len(nrow(esoph)), function(i) {
  with(esoph[i, ], data.frame(
    case  = c(rep(1L, ncases), rep(0L, ncontrols)),
    agegp = agegp, alcgp = alcgp, tobgp = tobgp
  ))
}))
# Use alcohol as an unordered factor so we get one OR per consumption band.
esoph_long$alcgp <- factor(esoph_long$alcgp, ordered = FALSE)

fit_alc <- matcha(
  esoph_long,
  outcome = "case", exposure = "alcgp",
  design = unmatched_cc(),
  confounders = ~ agegp + tobgp, estimator = "logistic"
)
contrast(fit_alc, type = "or")
```

A factor exposure reports one OR per non-reference level versus the baseline
(here `0-39g/day`), recovering the canonical monotone alcohol dose-response.
The reference level used is recorded on the result:

```{r}
fit_alc_result <- contrast(fit_alc, type = "or")
fit_alc_result$reference
```

### An ordinal trend

Entering an ordered exposure as a numeric *score* gives a single per-step trend
OR instead of per-level contrasts:

```{r}
esoph_long$alc_score <- as.integer(factor(
  esoph_long$alcgp,
  levels = c("0-39g/day", "40-79", "80-119", "120+")
))

fit_trend <- matcha(
  esoph_long,
  outcome = "case", exposure = "alc_score",
  design = unmatched_cc(),
  confounders = ~ agegp + tobgp, estimator = "logistic"
)
contrast(fit_trend, type = "or")
```

### Smooth confounder adjustment with `model_fn`

The logistic fitter is pluggable. Pass `model_fn = mgcv::gam` and a smooth term
in `confounders` to adjust for a confounder flexibly while keeping the exposure
parametric (so it still has an interpretable OR). The smooth-basis coefficients
are not reported as odds ratios.

```{r}
#| eval: !expr requireNamespace("mgcv", quietly = TRUE)
fit_gam <- matcha(
  cc,
  outcome = "case", exposure = "x",
  design = unmatched_cc(),
  confounders = ~ s(age), model_fn = mgcv::gam, estimator = "logistic"
)
contrast(fit_gam, type = "or")
```

## Mantel-Haenszel stratified OR

When confounding is controlled by *stratification* rather than a model, the
Mantel-Haenszel estimator pools the per-stratum 2×2 odds ratios in closed form.
Declare the stratifying variable(s) on the design and select `estimator = "mh"`.

```{r}
set.seed(3)
n <- 8000
agegrp <- sample(c("30s", "40s", "50s", "60s"), n, replace = TRUE)
x <- rbinom(n, 1, 0.4)
# True OR for x is 3, with an age-group main effect (confounding).
lp <- -1 + log(3) * x + 0.2 * as.integer(factor(agegrp))
case <- rbinom(n, 1, plogis(lp))
d_mh <- data.frame(case = case, x = x, agegrp = agegrp)

fit_mh <- matcha(
  d_mh,
  outcome = "case", exposure = "x",
  design = unmatched_cc(strata = "agegrp"), estimator = "mh"
)
contrast(fit_mh, type = "or")
```

The reported interval is the Robins-Breslow-Greenland (1986) variance, valid in
both the sparse-data and large-stratum limits. The MH estimator needs a 2×2
table per stratum, so the exposure must be binary; a categorical (k > 2) or
continuous exposure is routed to `estimator = "logistic"` with a classed error.
With no `strata`, `"mh"` reduces to the crude single-table OR.

## What the design does and does not identify

Only the odds ratio is identified from an unmatched case-control sample. Risk
differences and risk ratios require the source-population prevalence q0 and a
case-control-weighted estimator:

```{r}
#| error: true
contrast(fit, type = "difference")
```

## Covered combinations

**Legend.** ✅ truth-pinned in tests · ⛔ rejected with an informative error.

| Exposure | Estimator | Estimand | Variance | Status |
|---|---|---|---|---|
| Binary | `logistic` | conditional OR | model / sandwich | ✅ |
| Continuous | `logistic` | per-unit OR | model / sandwich | ✅ |
| Categorical (k>2) | `logistic` | per-level OR | model / sandwich | ✅ |
| Ordinal score | `logistic` | per-step trend OR | model / sandwich | ✅ |
| Any (GAM-adjusted) | `logistic` + `model_fn` | conditional OR | model | ✅ |
| Binary | `mh` | Mantel-Haenszel OR | Robins-Breslow-Greenland | ✅ |
| Ordered factor | `logistic` | — | — | ⛔ use a numeric score |
| Categorical / continuous | `mh` | — | — | ⛔ use `logistic` |
| Any | `logistic` / `mh` | RD / RR | — | ⛔ need q0 |

See `FEATURE_COVERAGE_MATRIX.md` for the authoritative status of every
combination.

## References

Breslow NE, Day NE (1980). *Statistical Methods in Cancer Research, Volume 1:
The Analysis of Case-Control Studies*. IARC.

Prentice RL, Pyke R (1979). Logistic disease incidence models and case-control
studies. *Biometrika* 66(3):403-411.

Robins J, Breslow N, Greenland S (1986). Estimators of the Mantel-Haenszel
variance consistent in both sparse data and large-strata limiting models.
*Biometrics* 42(2):311-323.
